### 2023-08-05 reports/paper/paper.qmd
New intervention

**Keywords:** Metalworking fluid, occupational exposure, non-Hodgkin
lymphoma, cancer, positivity, overlap, causal inference, g-formula,
iterated conditional expectation, healthy-worker effect

**Word count** **(Limit: 1500 to 4000 words):** 3812

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased
substantially in the US since 1960 and was recently linked to workplace
exposure to soluble metalworking fluids (MWF) in a standard survival
analysis of the United Auto Workers-General Motors (UAW-GM) cohort.
Here, we estimate the effects of hypothetical interventions on soluble
MWF exposure in relation to NHL risk in the same cohort of autoworkers.
Estimation of causal effects requires positivity, which we guarantee by
investigating causal contrasts that are supportable by the data.

**Methods:** We estimated counterfactual risk of NHL between 1985 and
2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n =
33,134) under supportable interventions on exposure to soluble MWF. The
supportable intervention reduces annual daily average exposures above
the target limit to that limit when the target limit is supported by
data. Otherwise, the supportable intervention reduces exposure to the
highest level of exposure below the target limit or not at all,
depending on support.

@@

**Conclusion:** Stronger limits on exposure to soluble MWF provide
stronger protections against NHL. We expect the effects of enforcing the
target exposure limits uniformly to have even stronger protective
effects. Strengthening exposure limits for MWF may protect workers from
NHL during the anticipated boom in domestic manufacturing.

@@

Metalworking fluids are complex mixtures of oil, water, and chemical
additives that optimize metal machining operations. There are three
general types of MWF: straight, soluble, and synthetic. During shaping,
grinding, and cutting operations, MWFs are misted, poured, or blasted at
high pressure onto work surfaces to remove debris, cool metal, improve
efficiency, and prevent deterioration of tools. Although MWFs are
essential to manufacturing processes, they also present a potential
health hazard to exposed workers through inhalation or ingestion of MWF
particulate mass. In response to health concerns and the rising cost of
oil, soluble MWF were developed to replace straight MWF and are now the
most commonly used type [@Byers_2006].

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects of exposure leave work [@Arrighi_1994]. An early causal
inference method in statistics developed to estimate causal effects in
longitudinal observational studies where the HWSE or other forms of
time-varying confounding/selection bias affected by past exposure may be
operating is the parametric g-formula [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal inference
is positivity (overlap) ie adequate variation in the exposure of
interest within strata formed by confounder and exposure histories
[@Petersen_2012].

Common approaches for addressing violations in positivity are
model-based smoothing when using outcome-regression techniques, weight
truncation when using propensity methods, or a combination of both.
Another approach is to consider only those causal contrasts that are
supported by the observed data ie have strictly positive propensity
score. Here, we defined supportable interventions on exposure to soluble
MWF in the United Auto Workers-General Motors (UAW-GM) occupational
cohort whose effects can be estimated with guaranteed positivity. We
estimated the cumulative incidence of NHL between 1985 and 2015 under
supportable interventions based on selected target exposure limits on
annual average daily exposure by applying the the hazard-extended
iterative conditional expectation (ICE) parametric g-formula
[@Arrighi_1994; @Wen_2020].

@@

We estimated NHL cumulative incidence from 1985 to 2015 under
supportable interventions on on annual average daily exposure to soluble
MWF based on several target exposure limits. First, we estimated the
expected number of NHL cases that we would observe if there were no
censoring by censoring and no target exposure limit. Then, we contrasted
this case count to that under supportable interventions based on five
hypothetical target exposure limits and no censoring. The five target
exposure limits were (1) 2.0, (2) 1.0, (3) 0.5, (4) 0.25, and (5) 0.05
mg/m^3^. The National Institute for Occupational Safety and Health
(NIOSH) recommended exposure limit (REL) for time-weighted average total
particulate mass (PM) composed of MWF is 0.5 mg/m^3^ [@niosh_1998].

For each target exposure limit, period of follow-up, and strata defined
by unique combinations of confounder and exposure histories, we found a
supportable exposure limit, which was the maximum observed value of
exposure at or below the target exposure limit. If all of the observed
values were above the target limit, no limit was enforced for that
stratum. The supportable intervention rule then reduces exposures above
the supportable exposure limit to that limit, but allowed exposures at
or below the limit to vary according to their observed distribution.
Applying the supportable intervention rule to the observed distribution
of exposure produces the intervention distribution that defines the
corresponding stochastic dynamic intervention with guaranteed
positivity. We estimated the effect of supportable intervention rules
based on selected target exposure limits, expressed as stochastic
dynamic interventions, using the hazard-extended ICE parametric
g-formula [@Wen_2020].

@@

We used data from the UAW-GM cohort, which included all hourly workers
at three automobile manufacturing plants in Michigan who had worked at
least three years by 1985. Past papers provide detailed descriptions of
the cohort [@Eisen_1992; @Eisen_2001]. The large size of the study
population and rich time-varying, quantitative MWF exposure data enable
the study of this relatively rare cancer and evaluate realistic
interventions on MWF exposure in a longitudinal cohort setting. The
present study population (N = 33,134) was restricted to the autoworkers
who were at work in 1941 or not yet hired, missing no more than half of
their employment history, and still alive at the start of follow-up.
Autoworkers in the study population were followed for NHL incidence from
January 1, 1985 until NHL diagnosis, death, December 31, 2014 (30 years
after the end of employment record availability), or the oldest observed
age at death (108 years), whichever came earlier.

@@

Potential confounders including year of hire, sex, race, time off,
employment status, and plant location were obtained from company
records. Race was missing for about 16% of the cohort, most commonly
among workers hired before 1960 in Plant 2. In analyses, missing race
was considered a distinct category. All potential confounders were coded
as categorical variables. Cut-points for categorizing continuous
covariates were determined according to the quantiles among cases.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
occupational cohort studies. Exposure assessment was based on direct air
sampling as well as company records. Company industrial hygienists
collected several hundred personal and area samples for total
particulate matter (mg/m^3^) composed of MWF over many decades. Research
industrial hygienists collected additional air sampling data when the
cohort study was launched in the mid 1980s. These additional data
combined with historical data and company records constituted a
job-exposure matrix of quantitative 8-hour time-weighted average daily
exposure estimates to soluble, straight, and synthetic MWF for each
combination of job, department, and plant over time. Workers' annual
average daily exposure to each MWF type was determined by combining this
job-exposure matrix with employment records, which recorded time-varying
job type, department, and plant for each employee from hire to
termination or 1995, whichever came sooner. For employment records that
were at least half complete, gaps in the record were interpolated by
carrying forward the last known job type. The exposure assessment is
described in detail elsewhere [@Hallock_1994; @Woskie_1994;
@Woskie_2003].

Previous analyses of NHL applied exposure lags of 1 to 20 years to
account for disease latency; we lagged cumulative MWF exposures by 20
years and therefore ended follow-up on January 1, 2015 [@Smith_2007;
@Karipidis_2007; @Zhang_2019]. In analyses, MWF exposure history at
start of follow-up was summarized as the cumulative sum of average
exposure intensities. Exposure was coded as categorical variables with
cut-points at zero and the quintiles of nonzero exposure among cases. We
estimated the effects of interventions on soluble MWF, the type of MWF
used most widely and in the greatest quantities while treating
co-exposure to straight and synthetic MWFs as potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the
deciles of the dates of diagnosis of NHL. The number of years per period
ranged from two to four years. The hazard-extended ICE parametric
g-formula involves two stages. In the first, we estimate counterfactual
discrete hazards over the person-periods. In the second, we pool those
estimates to estimate the counterfactual risk over the entire follow-up
period. During pooling, we iteratively combine estimates of the
counterfactual hazard to obtain a pooled estimate over an increasing
number of periods starting from the last period. In each iteration, we
perform model-based standardization over exposure and covariate
histories before combining the counterfactual discrete hazard estimate
with the estimated hazard pooled over subsequent periods. This iterative
process results in a sequentially standardized estimate of the
counterfactual cumulative incidence of NHL when the intervention of
interest is enforced over all follow-up periods.

We investigated supportable interventions that guarantee positivity:
every value of exposure that could be assigned under the our stochastic
dynamic intervention has a strictly positive propensity score. We took
exposure to straight and synthetic MWF, employment status, cumulative
time off, age, duration of employment, sex (male/female), race
(Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3) to be
potential confounders. Exposure to MWF, employment status, time off, and
duration of employment were lagged 20 years. An overview of the general
steps of the estimation procedure are presented below. Note that we
refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression of NHL on potential confounders and
    exposure over all at-risk person-periods.

@@

We estimated cumulative incidence under no intervention on soluble MWF
and under the supportable interventions based on the five target
exposure limits. We contrasted the cumulative incidence under the
supportable interventions to that under no intervention on exposure by
computing the number of cases averted and the cumulative incidence
ratios. Confidence intervals were computed using the basic nonparametric
bootstrap with 1000 Monte Carlo samples from the population at the start
of follow-up. All the necessary script used to reproduce the analyses
are available on [GitHub](https://github.com/kvntchn/gm-nhl-ice).

@@

Table 1 presents summary statistics for the full study population and
among those diagnosed with NHL between 1985 and 2015. The cohort is
predominantly white (64%) and male (86%). Year of hire and age at hire
were approximately the same among those with NHL and the full study
population. Median lagged cumulative exposure to all three types of MWF
was higher among NHL cases. Soluble MWF were the most widely used type,
with approximately 88% of workers ever exposed. Median cumulative
exposure among the exposed was 6 times higher for soluble than for
straight MWF. Figure 2 shows median annual average daily exposure to the
three MWF types among exposed workers over calendar time. Exposure to
MWF generally followed a downward trend over time.

@@

```{r}
tab1.sub |> 
  data.frame() |> flextable() |>
  delete_part(part = "header") |>
  add_header_row(values = c("", "Study population", "", "NHL cases"),
                 colwidths = c(1, 2, 1, 2)) |>
  hline_top(part = "header") |>
  hline_top() |> hline_bottom() |>
  hline(which((grepl("[0-9]\\)", tab1.sub$spread))[-1])[1]) |> width(4, 0.1) |>
  width(1, 2.2) |> width(c(3, 6), 1.1) |> width(c(2, 5), 0.65) |>
  flextable::compose(
    i = grep("Cumulative exposure", rownames(tab1.sub)), j = 1,
    value = as_paragraph("Cumulative exposure to MWF (mg/m",
                         as_sup("3"), "-years)")) |>
  flextable::align(j = 2:6, align = "center", part = "all") |>
  footnote(i = grep("Plant$", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Plant of longest employment duration among those who worked at multiple plants"),
    ref_symbols = "a", inline = T) |>
  footnote(i = grep("Ever exposed|time off", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Lagged 10 years"),
    ref_symbols = "b", inline = T) |>
  footnote(i = grep("work", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Among those who left work by December 31, 1994"),
    ref_symbols = "c", inline = T) |>
  # footnote(i = grep("death", rownames(tab1.sub)), j = 1, value = as_paragraph(
  #   " Among deceased"),
  #   ref_symbols = "d", inline = T) |>
  footnote(i = grep("Cumulative exposure", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Among ever-exposed individuals, lagged 10 years."),
    ref_symbols = "d", inline = T) |>
  add_footer_row(values = "NHL: non-Hodgkin lymphoma.", colwidths = 6) |>
  padding(padding = 1.5, part = "all") |>
  padding(
    i = which(tab1.sub_hspace),
    j = 1,
    padding.left = 20
  ) |>
  merge_at(i = grep("Cumulative exposure", rownames(tab1.sub))) |>
  hline_bottom(part = "footer") |> font(fontname = 'arial') |>
	set_caption(
		caption = paste0(
			"Summary of population characteristics. Statistics shown above ",
			"the horizontal line are count (%). Those shown below are median ",
			"(quartile 1, quartile 3).")
	)
```

@@

The observed number of NHL cases over the 30-year follow-up period was
339. Table 2 presents the hazard-extended ICE parametric g-formula
estimates of the counterfactual number of cases, number of cases
averted, and cumulative incidence ratios contrasting supportable
interventions on exposure to soluble MWF based on different target
exposure limits and no censoring. Under an intervention eliminating
censoring, the estimated number of cases was 502 (95% CI: 439, 555).
Interventions based on stronger target exposure limits on soluble MWF
resulted in monotonically stronger reductions in the estimated
cumulative incidence of NHL. Setting the target exposure limit at the
NIOSH REL 0.5 mg/m^3^ would have averted 112 (95% CI: 57, 181) NHL
cases.

@@

Although NIOSH concluded that there exists substantial evidence linking
MWF exposure to several different cancers including larynx, rectum,
pancreas, skin, scrotum, and bladder cancer, their REL of 0.5 mg/m^3^
for total particulate matter derived from any type of MWF targeted risk
of nonmalignant health conditions [@niosh_1998; @Mirer_2003]. Exposure
limits stronger than the NIOSH REL may provide valuable health
protections not previously considered by policy makers. Using the
hazard-extended ICE parametric g-formula, we estimated the
counterfactual expected number of NHL cases from 1985 to 2015 in the
UAW-GM cohort if we enforced supportable interventions on soluble MWF
based on five different target exposure limits and found a monotonic
exposure-dependent relationship where stronger target exposure limits
yielded lower NHL case count estimates.

The g-formula is a well-known approach in causal inference used for
estimating causal effects in the presence of time-varying confounding
affected by past exposure [@Robins_1986]. Standard representations of
the g-formula include (1) a non-iterated expectation over the joint
density of covariates, (2) the ICE over time, and (3) an inverse
probability weighted expectation. The classic parametric g-formula is a
plug in estimator for the g-formula under its first, non-iterative,
representation. It involves the parametric modeling of the full joint
distribution of the confounders, exposure, and outcome for each time
point [@Taubman_2009; @Young_2014]. Counterfactual quantities under
hypothetical interventions of interest are computed from Monte Carlo
samples from distributions implied by the fitted parametric models. In
longitudinal settings, this approach often requires specifying and
fitting large number of models in order to satisfy the exchangeability
assumptions necessary for causal identification. Despite this
limitation, analysts often choose the parametric g-formula because of
the intuitive way it handles interventions on the natural value of
exposure. However, these causal estimands are not unique to the
parametric g-formula. The distribution of exposure produced by
marginalizing intervention rules over the observed distribution of
exposure within strata formed by potential confounder and exposure
histories defines a corresponding stochastic dynamic intervention, whose
effects may be estimated using various estimators [@Wen_2020;
@Bang_2005; @van-der-Laan_2011; @Wen_2023; @Diaz_2021]. Estimating
causal effects of this implied stochastic dynamic intervention is
analogous to that of interventions on the natural value of exposure
[@Young_2014].
<!-- Since applications of the parametric g-formula are often context-laden, there are no unified guidelines for model specification. -->

@@

Positivity refers to the need for adequate variation in future exposure
among strata formed by observed covariate and intervention-compliant
exposure histories. Even under conditional exchangeability, where
exposures within these strata may be considered the result of
experimental assignment, expected counterfactual outcomes under
different exposures may not be estimable if there is excessive sparsity
in the observed distribution of exposures [@Maldonado_2002]. Rather than
address potential violations in positivity analytically, we investigated
stochastic dynamic interventions on soluble MWF exposure based on
intervention distributions which are nonzero only where the propensity
score is strictly positive. Hence, our supportable interventions
guarantee positivity. Interventions that guarantee positivity have been
suggested in the past, and have also been called "realistic"
interventions [@Petersen_2012].

@@

Much of the existing epidemiologic literature linking occupational and
environmental exposures to NHL report findings from case-control studies
where exposures are measured as binary indicators of exposure or as
membership in a particular occupational group [@Cano_2001;
@Costantini_2001; @Karunanayake_2008; @t-Mannetje_2016]. Associations
between occupations and NHL risk vary considerably, but one study of
working men in Kansas and Nebraska found strong associations between NHL
risk and occupations involving metalworking and motor vehicles
[@Zheng_2002]. Both of these occupations may entail exposure to soluble
MWF, which contain a number of additives of concern for human health and
for NHL risk in particular. Organic compounds containing phosphorous,
chlorine, sulfur, nitrogen, and boron are commonly added to soluble MWF
to control microbial growth, improve performance under high
heat/pressure, and inhibit corrosion [@Evans_2020]. Organophosphorus
compounds include organophosphate pesticides, which have been linked to
cancer risk in epidemiologic and animal studies. Some were classified as
possibly carcinogenic by the IARC [@IARC_2017]. Studies of occupational
exposure to chlorinated solvents and pesticides have also been linked to
NHL risk [@Cocco_2008; @Purdue_2011; @Cocco_2013; @Vlaanderen_2013;
@Callahan_2018]. In 2014, the IARC classified trichloroethylene,
tetrachloroethylene, and other chlorinated agents as Group 1 carcinogens
[@IARC_2014]. Chlorinated solvents are commonly used as degreasers in
industrial settings, but their use in the plants under study here was
rare [@Shrestha_2016]. The structural characteristics shared by MWF
additives and known/suspected carcinogens suggest potential similarities
in their behavior in biological systems.

>>>>

**Keywords:** Positivity, overlap, causal inference, g-formula,
metalworking fluid, occupational exposure, non-Hodgkin lymphoma, cancer,
iterated conditional expectation, healthy-worker effect

**Word count** **(Limit: 1500 to 4000 words):** 3812 (approx)

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased
substantially in the US and was recently linked to workplace exposure to
soluble metalworking fluids (MWF) in a standard survival analysis of the
United Auto Workers-General Motors (UAW-GM) cohort. Here, we estimate
the effects of hypothetical interventions on soluble MWF exposure in
relation to NHL risk in the same cohort of autoworkers. Estimation of
causal effects requires positivity, which we guarantee by investigating
causal contrasts that are supportable by the data.

**Methods:** We estimated counterfactual risk of NHL between 1985 and
2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n =
33,134) under hypothetical supportable interventions on exposure to
soluble MWF. A supportable intervention reduces annual daily average
exposures above a specific target limit to that limit when supported by
data. Otherwise, the supportable intervention reduces exposure to the
highest level below the target limit or not at all, depending on
support.

@@

**Conclusion:** More stringent limits on exposure to soluble MWF provide
stronger protections against NHL. We expect the effects of enforcing the
target exposure limits uniformly to have even stronger protective
effects. Strengthening exposure limits for MWF may protect workers from
NHL during the anticipated boom in domestic manufacturing.

@@

Metalworking fluids are complex mixtures of oil, water, and chemical
additives that cool and lubricate metal machining operations. There are
three general types of MWF: straight, soluble, and synthetic. During
shaping, grinding, and cutting operations, MWFs are misted, poured, or
blasted at high pressure onto work surfaces to remove debris, cool
metal, improve efficiency, and prevent deterioration of tools. Although
MWFs are essential to manufacturing processes, they also present a
potential health hazard to exposed workers through inhalation or
ingestion of MWF particulate mass. In response to health concerns
related to the carcinogenicity of polycyclic aromatic hydrocarbons
(PAHs) in mineral oil as well as the rising global cost of oil products,
water-based soluble MWF, which usually contain about 40-70% oil by
weight, were developed to replace oil-based straight MWF [@IARC_1973;
@IARC_1987; @Childers_2006]. Today, soluble MWF is the most common type
of MWF in metal machining operations [@Byers_2006].

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects of exposure leave work [@Arrighi_1994]. An early causal
inference method in statistics developed to estimate causal effects in
longitudinal observational studies where the HWSE or other forms of
time-varying confounding/selection bias affected by past exposure may be
operating is the parametric g-formula [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal inference
from observational data is positivity (overlap) ie adequate variation in
the exposure of interest within strata formed by confounder and exposure
histories [@Petersen_2012].

Common approaches for addressing violations in positivity are
model-based smoothing when using outcome regression methods, weight
truncation when using propensity methods, or a combination of both.
Another approach is to only consider causal contrasts that are supported
by the observed data ie those involving hypothetical interventions with
propensity score strictly greater than zero. Here, we defined
hypothetical supportable interventions on exposure to soluble MWF in the
United Auto Workers-General Motors (UAW-GM) occupational cohort and
estimated their effects on NHL incidence between 1985 and 2015 relative
to no exposure intervention while guaranteeing positivity.

@@

We estimated the cumulative incidence of NHL between 1985 and 2015 under
supportable interventions based on selected target exposure limits on
annual average daily exposure to soluble MWF by applying the the
hazard-extended iterative conditional expectation (ICE) parametric
g-formula [@Wen_2020]. First, we estimated the expected number of NHL
cases that we would observe if there were no censoring and no limit on
exposure. Then, we contrasted this case count to that under supportable
interventions based on five hypothetical target exposure limits and no
censoring. The five target exposure limits were (1) 2.0, (2) 1.0, (3)
0.5, (4) 0.25, and (5) 0.05 mg/m^3^. The National Institute for
Occupational Safety and Health (NIOSH) recommended exposure limit (REL)
for time-weighted average total particulate mass (PM) composed of MWF is
0.5 mg/m^3^ [@niosh_1998].

For each target exposure limit, period of follow-up, and stratum defined
by unique combinations of confounder and exposure histories, we found a
supportable exposure limit, which was the maximum observed value of
exposure at or below the target exposure limit, if such a value exists.
If all of the observed values were above the target limit, the
supportable limit was simply the maximum observed value. The supportable
intervention rule then reduces exposures above the supportable exposure
limit to that limit, but allows exposures at or below the limit to vary
according to their observed distribution. Applying the supportable
intervention rule to the observed distribution of exposure produces the
intervention distribution that defines the corresponding stochastic
dynamic intervention with guaranteed positivity. We estimated the effect
of supportable intervention rules based on the selected target exposure
limits, expressed as stochastic dynamic interventions, using the
hazard-extended ICE parametric g-formula [@Wen_2020].

@@

We used data from the UAW-GM cohort study, which included all hourly
workers at three automobile manufacturing plants in Michigan who had
worked at least three years by 1985. Past papers provide detailed
descriptions of the cohort [@Eisen_1992; @Eisen_2001]. The large size of
the study population and rich time-varying, quantitative MWF exposure
data enable the study of a relatively rare cancer and evaluate realistic
interventions on MWF exposure in a longitudinal cohort setting. The
present study population (N = 33,134) was restricted to the autoworkers
who were at work in 1941 or not yet hired, missing no more than half of
their employment history, and still alive at the start of follow-up for
cancer incidence on January 1, 1941. Autoworkers in the study population
were followed until NHL diagnosis, death, December 31, 2014, or the
oldest observed age at death, whichever came earlier.

@@

Potential confounders including year of hire, sex, race, time off work,
employment status, and plant location were obtained from company
records. Race was missing for about 16% of the cohort, most commonly
among workers hired before 1960 in Plant 2. In analyses, missing race
was considered a distinct category. All potential confounders were coded
as categorical variables. Cut-points for categorizing continuous
covariates were determined according to the quantiles among cases.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
environmental or occupational health studies. Exposure assessment was
based on direct air sampling as well as company records. Company
industrial hygienists collected several hundred personal and area
samples for total particulate mass (mg/m^3^) composed of MWF over many
decades. Research industrial hygienists collected additional air
sampling data when the cohort study was launched in the mid 1980s
[@Woskie_1994; @Woskie_1994a]. These additional data combined with
historical data and company records constituted a job-exposure matrix of
quantitative 8-hour time-weighted average daily exposure estimates to
soluble, straight, and synthetic MWF for each combination of job,
department, and plant over time [@Hallock_1994]. Workers' annual average
daily exposure to each MWF type was determined by combining this
job-exposure matrix with employment records, which recorded time-varying
job type, department, and plant for each employee from hire to
termination or 1995, whichever came sooner. For employment records that
were at least half complete, gaps in the record were interpolated by
carrying forward the last known job type. The exposure assessment is
described in detail elsewhere [@Hallock_1994; @Woskie_1994;
@Woskie_2003].

In previous analyses of NHL, exposure lags of 1 to 20 years were applied
to account for disease latency; we lagged cumulative MWF exposures by 20
years [@Smith_2007; @Karipidis_2007; @Zhang_2019]. In analyses, MWF
exposure history at start of follow-up was summarized as the cumulative
sum of average exposure intensities. Exposure was coded as categorical
variables with cut-points at zero and the quintiles of nonzero exposure
among cases. We estimated the effects of hypothetical interventions on
soluble MWF, the type of MWF used most widely and in the greatest
quantities while treating co-exposure to straight and synthetic MWFs as
potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the
deciles of the dates of diagnosis of NHL. The number of years per period
ranged from two to four. The hazard-extended ICE parametric g-formula
involves two stages. In the first, we estimate counterfactual discrete
hazards over the person-periods. In the second, we pool those estimates
to estimate the counterfactual risk over the entire follow-up period.
During pooling, we iteratively combine estimates of the counterfactual
hazard to obtain a pooled estimate over an increasing number of periods
starting from the last period. In each iteration, we perform model-based
standardization over exposure and covariate histories before combining
the counterfactual discrete hazard estimate with the estimated hazard
pooled over subsequent periods. This iterative process results in a
sequentially standardized estimate of the counterfactual cumulative
incidence of NHL when the intervention of interest is enforced over all
follow-up periods.

We investigated supportable interventions that guarantee positivity:
every value of exposure that could be assigned under the our stochastic
dynamic intervention has propensity score strictly greater than zero. We
treated co-exposure to straight and synthetic MWF, employment status,
cumulative time off, age, duration of employment, sex (male/female),
race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3) as
potential confounders. Exposure to MWF, employment status, time off, and
duration of employment were lagged 20 years. An overview of the general
steps of the estimation procedure are presented below. Note that we
refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression model for NHL on potential
    confounders and exposure over all at-risk person-periods.

@@

We estimated cumulative incidence under no intervention on soluble MWF
and under the supportable interventions based on the five selected
target exposure limits. We contrasted the cumulative incidence under the
supportable interventions to that under no intervention on exposure by
computing the number of cases averted and the cumulative incidence
ratios. Confidence intervals were computed using the basic nonparametric
bootstrap with 1000 Monte Carlo samples from the population at the start
of follow-up. All the necessary scripts used to reproduce the analyses
are available on [GitHub](https://github.com/kvntchn/gm-nhl-ice).

@@

Table 1 presents summary statistics for the full study population and
among those diagnosed with NHL between 1985 and 2015. The cohort is
predominantly white (64%) and male (86%). Year of hire and age at hire
were approximately the same among those with NHL and the full study
population. Median lagged cumulative exposure to all three types of MWF
was higher among NHL cases. Soluble MWF were the most widely used type,
with approximately 88% of workers ever exposed. Median cumulative
exposure among the exposed was approximately 6 times higher for soluble
than for straight MWF. Figure 2 shows median annual average daily
exposure to the three MWF types among exposed workers over calendar
time. Exposure to MWF generally followed a downward trend over time.

@@

```{r}
tab1.sub |> 
  data.frame() |> flextable() |>
  delete_part(part = "header") |>
  add_header_row(values = c("", "Study population", "", "NHL cases"),
                 colwidths = c(1, 2, 1, 2)) |>
  hline_top(part = "header") |>
  hline_top() |> hline_bottom() |>
  hline(which((grepl("[0-9]\\)", tab1.sub$spread))[-1])[1]) |> width(4, 0.1) |>
  width(1, 2.2) |> width(c(3, 6), 1.1) |> width(c(2, 5), 0.65) |>
  flextable::compose(
    i = grep("Cumulative exposure", rownames(tab1.sub)), j = 1,
    value = as_paragraph("Cumulative exposure to MWF (mg/m",
                         as_sup("3"), "-years)")) |>
  flextable::align(j = 2:6, align = "center", part = "all") |>
  footnote(i = grep("Plant$", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Plant of longest employment duration among those who worked at multiple plants"),
    ref_symbols = "a", inline = T) |>
  footnote(i = grep("Ever exposed|time off", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Lagged 20 years"),
    ref_symbols = "b", inline = T) |>
  # footnote(i = grep("work", rownames(tab1.sub)), j = 1, value = as_paragraph(
  #   " Among those who left work by December 31, 1994"),
  #   ref_symbols = "c", inline = T) |>
  # footnote(i = grep("death", rownames(tab1.sub)), j = 1, value = as_paragraph(
  #   " Among deceased"),
  #   ref_symbols = "d", inline = T) |>
  footnote(i = grep("Cumulative exposure", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Among ever-exposed individuals, lagged 20 years."),
    ref_symbols = "d", inline = T) |>
  add_footer_row(values = "NHL: non-Hodgkin lymphoma.", colwidths = 6) |>
  padding(padding = 1.5, part = "all") |>
  padding(
    i = which(tab1.sub_hspace),
    j = 1,
    padding.left = 20
  ) |>
  merge_at(i = grep("Cumulative exposure", rownames(tab1.sub))) |>
  hline_bottom(part = "footer") |> font(fontname = 'arial') |>
	set_caption(
		caption = paste0(
			"Summary of population characteristics. Statistics shown above ",
			"the horizontal line are count (%). Those shown below are median ",
			"(quartile 1, quartile 3).")
	)
```

@@

The observed number of NHL cases over the 30-year follow-up period was
339. Table 2 presents estimates of the counterfactual number of cases,
number of cases averted, and cumulative incidence ratios contrasting
supportable interventions on exposure to soluble MWF based on different
target exposure limits and no censoring. Under an intervention
eliminating censoring, the estimated number of cases was 502 (95% CI:
439, 555). Interventions based on stronger target exposure limits on
soluble MWF resulted in monotonically stronger reductions in the
estimated cumulative incidence of NHL. Setting the target exposure limit
at the NIOSH REL 0.5 mg/m^3^ would have averted 112 (95% CI: 57, 181)
NHL cases.

@@

Although NIOSH concluded that there exists substantial evidence linking
MWF exposure to several different cancers including larynx, rectum,
pancreas, skin, scrotum, and bladder cancer, their REL of 0.5 mg/m^3^
for total particulate mass derived from any type of MWF was based on
nonmalignant respiratory health effects [@niosh_1998; @Mirer_2003].
Exposure limits stronger than the NIOSH REL may provide valuable health
protections not previously taken into account by policy makers. Using
the hazard-extended ICE parametric g-formula, we estimated the
counterfactual expected number of NHL cases from 1985 to 2015 in the
UAW-GM cohort if we enforced hypothetical supportable interventions on
soluble MWF based on five different target exposure limits and found a
monotonic exposure-dependent relationship where stronger target exposure
limits yielded lower NHL case count estimates.

The g-formula is a well-known approach in causal inference used for
estimating causal effects in the presence of time-varying confounding
affected by past exposure [@Robins_1986]. Standard representations of
the g-formula include (1) a non-iterated expectation over the joint
density of covariates, (2) the ICE over time, and (3) an inverse
probability weighted expectation. The classic parametric g-formula is a
plug-in estimator for the g-formula under its first, non-iterative,
representation. It involves the parametric modeling of the full joint
distribution of the confounders, exposure, and outcome for each time
point [@Taubman_2009; @Young_2014]. Counterfactual quantities under
hypothetical interventions of interest are computed from Monte Carlo
samples from distributions implied by the fitted parametric models. In
longitudinal settings, this approach often requires specifying and
fitting large number of models in order to satisfy the exchangeability
assumptions necessary for causal identification. Despite this
limitation, analysts often choose the parametric g-formula because of
the intuitive way it handles interventions on the natural value of
exposure. However, these causal estimands are not unique to the
parametric g-formula. The distribution of exposure produced by
marginalizing intervention rules over the observed distribution of
exposure within strata formed by potential confounder and exposure
histories defines a corresponding stochastic dynamic intervention, whose
effects may be estimated using various estimators [@Wen_2020;
@Bang_2005; @van-der-Laan_2011; @Wen_2023; @Diaz_2021]. Estimating
causal effects of this implied stochastic dynamic intervention is
analogous to that of interventions on the natural value of exposure
[@Young_2014].
<!-- Since applications of the parametric g-formula are often context-laden, there are no unified guidelines for model specification. -->

@@

Positivity refers to the need for adequate variation in future exposure
among strata formed by observed covariate and intervention-compliant
exposure histories. Even under conditional exchangeability, where
exposures within these strata may be considered the result of
experimental assignment, expected counterfactual outcomes under
different exposures may not be estimable if there is excessive sparsity
in the observed distribution of exposures [@Maldonado_2002]. Rather than
address potential violations in positivity analytically, we investigated
stochastic dynamic interventions on soluble MWF exposure based on
intervention distributions which are nonzero only where the propensity
score is strictly positive. Hence, our supportable interventions
guarantee positivity. Interventions that guarantee positivity have been
suggested in the past and have also been called "realistic"
interventions [@Petersen_2012].

@@

Much of the existing epidemiologic literature linking occupational and
environmental exposures to NHL report findings from case-control studies
where exposures are measured as binary indicators of exposure or as
membership in a particular occupational group [@Cano_2001;
@Costantini_2001; @Karunanayake_2008; @t-Mannetje_2016]. Associations
between occupations and NHL risk vary considerably, but one study of
working men in Kansas and Nebraska found strong associations between NHL
risk and occupations involving metalworking and motor vehicles
[@Zheng_2002]. Both of these occupations may entail exposure to soluble
MWF, which contain mineral oils and a number of chemical additives of
concern for human health and for NHL risk in particular. Organic
compounds containing phosphorous, chlorine, sulfur, nitrogen, and boron
are commonly added to soluble MWF to control microbial growth, improve
performance under high heat/pressure, and inhibit corrosion
[@Evans_2020]. Organophosphorus compounds include organophosphate
pesticides, which have been linked to cancer risk in epidemiologic and
animal studies. Some were classified as possibly carcinogenic by the
IARC [@IARC_2017]. Studies of occupational exposure to chlorinated
solvents and pesticides have also been linked to NHL risk [@Cocco_2008;
@Purdue_2011; @Cocco_2013; @Vlaanderen_2013; @Callahan_2018]. In 2014,
the IARC classified trichloroethylene, tetrachloroethylene, and other
chlorinated agents as Group 1 carcinogens [@IARC_2014]. Chlorinated
solvents are commonly used as degreasers in industrial settings, but
their use in the plants under study here was rare [@Shrestha_2016]. The
structural characteristics shared by MWF additives and known/suspected
carcinogens suggest potential similarities in their behavior in
biological systems.
